Transfer learning for echelle spectroscopy

### 2022-02-22 paper/paper1/ms.tex
grammarly to the rescue

\subsection{Seeking physical self-consistency and high predictive accuracy}
Ideally we seek models that are both physically self-consistent and highly accurate when compared to data.  We want tunable models that can be evaluated quickly for use in likelihood calculations.  We want posterior inference to accurately explore correlations among parameters.  We want unbiased, noise-free templates for use in cross-correlation spectroscopy.  We want accuracy over a large spectral grasp: simultaneously wide bandwidth and high spectral resolution.  We want high out-of-band prediction accuracy.  Ideally we also want precision and accuracy in the separation of affiliated telluric and instrumental signals.

@@

\subsection{This paper: towards semi-empirical models}
\begin{itemize}
  \item So in practice, tradeoffs between model flexibility and physical self-consistency
  \item Semi-empirical models may be a middle ground: informed from models, but revised with data
  \item The radiative transfer step is computationally expensive, so we want to avoid redoing that
  \item Instead focus on pre-computed models that have that wisdom baked-in
  \item We are essentially compressing pre-computed models into evaluable, interpretable models
  \item Transfer learning is one plausible approach: pre-train on pre-computed models, transfer to data
  \item Neural network frameworks make this possible with autodiff
  \item We use the autodiff and GPU acceleration, but not the neural architectures
  \item We do not use the neural architectures because we have good understand for how a spectrum is generated
  \item Instead of tuning weights of a neural network, we tune the atomic and molecular properties of a spectrum
\end{itemize}

@@

We start with a high resolution pre-computed synthetic stellar model spectrum, $\mathsf{S}(\bm{\lambda})$.  The procedure is largely agnostic to the exact details of how this spectrum was made.  For the purposes of this paper, we will showcase examples from two well-known families of precomputed synthetic spectra: \texttt{PHOENIX} \citep{husser13} for stellar spectra $(T_{\mathrm{eff}}\in [2300, 10000]\;K)$ and \texttt{Sonora} \citep{2021ApJ...920...85M} for brown dwarf spectra $(T_{\mathrm{eff}}\in [1000, 2300]\;K)$.  We place the following demands on the the precomputed spectra.  They should have sporadic regions of discernable continuum devoid of lines, and the continuum must vary smoothly in wavelength.  The spectral lines or pseudo-lines should be resolved, and not sub-sampled.  We suspect most stellar spectra meet these criteria, with the exception of the coolest M dwarfs and brown dwarfs.  The method can hypothetically handle spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed.  We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars.  A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

The choice of limiting the bandwidth to a region-of-interest around a single echelle spectrograph bandwidth stems from computational constraints.  In principle there is no fundamental limit to the bandwidth one could clone with the method presented here, up to and including the entire precomputed synthetic spectral model bandwidth.  We adopt the exact native wavelength sampling with no smoothing or interpolation, yielding a wavelength vector $\bm{\lambda}_s$ with length $N_\mathrm{s}$ equal to the number of pixels within the extents of our region of interest including the buffers.

At this stage we have the choice of whether to work in linear or log scale flux units.  Adopting the log of the flux would ensure that the cloned model possesses only positive flux values, a desireable trait of any physical spectral model.  We chose instead to stick with native linear flux scaling because of the availability of convenient approximations to astrophysically relevant line profiles in the linear scaling domain.  We normalize the spectrum by its median value since we do not intend to apply these methods to absolutely flux calibrated spectra.

@@

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template.  The resulting signal typically possesses a smooth continuum variation.  An additional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly.  Here, we identify the continuum regions by the most prominent peaks in the specturm separated by a distance of at least 5,000$-$90,000 pixels from adjacent peaks, yielding between 60 and 5 control points.  We then fit an $n_{\mathrm{poly}} = 3^{rd}-15^{th}$ order polynomial to these peaks, evaluate the polynomial on the entire bandwidth $\mathsf{P}(\bm{\lambda}_s)$, and divide the spectrum by this trend.  Visual inspection of this continuum-flattened spectrum typically possesses variations at the $<1\%$ level.  Tuning the pixel distance and polynomial order can lower the undulations to $<0.05\%$ level for some input spectra.  We found a pixel distance of 50,000 and $3^{rd}$ order polynomial worked across the range of effective temperatures for HPF's bandwidth. The IGRINS spectra had more significant continuum opacity and bandheads to accomodate and therefore needed to capture finer scale variation: 5,000 pixels and a $9^{th}$ order polynomial appeared adequate over most of the effective temperature range.

This high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line-wings such as those in deep Hydrogen and sodium lines.  We have experimentally found it helpful to hand-tune models of exceptionally broad and non-standard line-wings with an optional pre-processing step that depends on the wavelength range of interest.  Here we sketch the approach for common broad lines found in the long-wavelength visible and near-infrared wavelength regions.

@@

We emphasize that a recreation of the unvarnished input spectrum can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda}_s)\cdot \mathsf{P}(\bm{\lambda}_s)$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low resolution spectra, or regions with molecular bandheads.  The ``perturbed black body'' continuum model contains $n_{\mathrm{poly}}+1$ fixed-but-possibly tunable lookup parameters, plus the fixed input $T_{\mathrm{eff}}$.

Next we identify the spectral lines.  We apply a local-minimum-finding algorithm on the spectrum by defining a prominence threshold $P_{rom} \in (0.005, 0.02)$.  This threshold dictates the number of lines that will be modeled: a lower prominence finds more, weaker lines, and a larger prominence finds fewer, deeper lines.  The prominence algorithm successfully finds lines that reside on top of broad line wings, or unresolved bandheads provided that the individual lines exceed the prominence threshold in their local region.  The number of lines $N_{\mathrm{lines}}$ depends on the bandwidth, prominence, and the intrinsic properties of the input spectrum, principally effective temperature and metallicity.

For this paper, we illustrate examples for two \'echelle spectrographs with particularly large spectral grasp: the Habitable Zone Planet Finder \citep[HPF,][]{2014SPIE.9147E..1GM} on the Hobby-Eberly Telescope at McDonald Observatory in Fort Davis, Texas; and the Immersion Grating Infrared Spectrograph \citep[IGRINS,][]{park14}
currently on the Gemini South Telescope on Cerro Pach\'on in Chile.  The $R=55,000$ HPF has a native bandwidth of $8079-12785$~\AA, that we expand to $8049-12815$~\AA~including the edge buffers.  IGRINS has two cameras for $H$ and $K$ band, with the combined spectrum spanning $14267-25217\;$\AA~ including the edge buffers and the region in-between the two cameras, all at a resolving power of $(R=45,000)$.  The spectrograph acquisition, reduction, and post-processing steps yield data $\mathsf{D}(\bm{\lambda}_{d})$, where $\bm{\lambda}_{d}$ is the wavelength vector at the instrumental resolution and sampling of each instrument, generally much coarser than the resolution and sampling grid of the precomputed synthetic spectra.  The data wavelength vector may also contain gaps between \'echelle orders, whereas the precomputed wavelength coordinates are usually contiguous.  HPF may have up to $2048\times28=$57,344 pixels, and IGRINS has typically about 75,000 pixels, after common trimming of noisy edge pixels and unusable telluric regions.

@@

Figure \ref{fig_Nlines_vs_teff} shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold $P_{rom}$ for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS.  We see between about 2,000 and 20,000 lines depending on the $T_{\mathrm{eff}}$ and $P_{rom}$.  HPF and IGRINS have comparable number of lines, and halving the prominence increases the number of lines by about $20-30\%$ in these ranges.  The number of lines monotonically increases towards cooler effective temperatures.
The HPF-truncated spectra have $N_s=$335,849 native resolution samples, comparable to the IGRINS-truncated spectra, $N_s=$330,052.

@@

There does not exist a general-purpose, single-shot algorithm for obtaining the lineshape in the presence of overlapping spectral lines: where do the wings of one line begin and the wings of another adjacent line end?  We therefore do not attempt to determine anything about the lineshape at this stage, and instead assume that the lines resemble a Voigt profile, with a guess width about equally split between Lorentzian and Gaussian.

@@

where $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ is the normalized Voigt profile with Gaussian standard deviation $\sigma$, Lorentzian half-width $\gamma$, at line center position $\lambda_c$, for the $j^{th}$ spectral line.  The amplitude $a$ is always expected to be positive for absorption lines, and is equal to the area under the curve of the line since the Voigt profile is normalized.  The optional Prefactor term $\mathsf{P}$ is fixed at $1$ by default, but can be allowed to vary to fine-adjust imperfections left over from the polynomial-based continuum flattening procedure.

\subsection{Voigt profile evaluation}
The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed form using the Fadeeva function\authorcomment1{citation}.  Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable.  Here we adopt the pseudo-Voigt approximation described in \censor{some paper}\authorcomment1{look up reference to PseudoVoigt}.

Two sampled Voigt approaches could be feasible.  Direct convolution of a sampled Gaussian and Lorentzian profile may be too costly and inexact if pixels are coarsely sampled.  Analogously, the convolution can be achieved in the Fourier domain through multiplication of the Fast Fourier Transforms of the Gaussian and Lorentzian profiles, followed by the subsequent Inverse Fourier Transforms.  The main rationale for either of these two choices would be to anticipate and align with planned convolutions such as rotational broadening and instrumental kernels.  We do not currently implement these sampled convolutions in \emph{blas\'e} due to their prohibitive computational cost.

\subsection{Goodness of fit metric}
The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes.  Instead, we tune the parameters of the model, starting from these coarse initial values.  This model has $N_{\mathrm{lines}}\times 3$ free parameters, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary.  We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{M}$ but has no notion of per-pixel noise, since the precomputed synthetic spectrum has no uncertainty:

@@

The optimizer updates the $a_j, \sigma_j, \gamma_j$ parameters by a small fraction of the Jacobian---called the learning rate LR---towards the direction that would improve the fit, for all $N_{\mathrm{lines}} \times 3$ parameters simultaneously.  The Jacobian is calculated behind-the-scenes with automatic differentiation\authorcomment1{Citation} implemented as the so-called backwards-propagation algorithm or simply ``back-prop''.  We choose the \texttt{PyTorch} framework that computes these Jacobians efficiently for all of the mathematical primatives in our \texttt{blas\'e} implementation.

In principle we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch.  Further, an autodiff-aware implementation of the \emph{celerit\'e} algorithm\authorcomment1{add DFM citation} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting.  In practice, PyTorch also offers hardware acceleration (on GPU or TPU)\authorcomment1{citation needed}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks.  The GPyTorch framework\authorcomment1{add citation} may offer a workaround to the absence of a \emph{celerit\'e} implementation.  Still, the performance of an analytic Jacobian implementation over the autodiff route presented here offers an interesting avenue of future research.

\subsection{GPU and Autodiff specific considerations}
We make a few tweaks to the implementation for numerical purposes.  First, we want all the parameters to be positive, forbidding negative amplitudes and negative widths.  We therefore tune the natural log of the parameters, and exponentiate them before inclusion in Equation \ref{equation1}.  Second, we found through iterative experimentation that the initialization amplitudes and widths were systematically shifted from the optimized values.  We built-in scalar tweaks to the initialization amplitudes, Gaussian width, and Lorentzian width, which dramatically accelerated the optimization step.

@@

Equation \ref{equation1} performs a type of tensor contraction, turning a $N_{\mathrm{lines}}\times N_{s}$ matrix into a $1\times N_{s}$ vector.  The number of Floating Point Operations (FLOPS) scales with the number of entries in this matrix, with a scalar prefactor for the cost of evaluating a single line profile at a single wavelength point.  Efficient GPU algorithms exist for tensor contractions such as this product along an axis of a matrix, allowing this computation to proceed quickly on modern machines\authorcomment1{add citation?}. In particular, the proprietary CUDA architecture for NVIDIA\textsuperscript{\tiny\textregistered} GPUs contains Tensor cores with specialized matrix math, although the algorithms place restrictions on floating precision depending on the particular hardware and software available.  The chief bottleneck occurs when the storage of the $\bm{\bar{V}}$ matrix exceeds the available RAM of a GPU or CPU: the computation will fail with an ``Out of Memory'' exception. Modern NVIDIA GPUs have $8-40$ GB of RAM.  The memory bottleneck is even more pernicious than mere storage of the $\bm{\bar{V}}$ matrix, since the CPU/GPU also has to store the computation graph for autodiff.  The computation graph takes up more memory than the matrix itself, so it is generally not possible to evaluate Equation \ref{equation1} in its entirety in one-fell-swoop.  A remedy is needed.

@@

Minibatches offer a natural solution to the line-by-line computational bottleneck while also providing optimization performance improvements.  Minibatches act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent\authorcomment1{citation}.  The Machine Learning and Artificial Intelligence (ML/AI) community employs minibatches to handle similar situations, in which the dataset volume exceeds the available GPU RAM.  Our configuration departs slightly from the typical ML/AI situation, since their models are usually ``small'' relative to their datasets.  Here our model is ``big'' relative to our dataset.

Here we choose to assemble and evaluate only a portion of the matrix at a time, in minibatches.  The choice to evaluate only a portion of lines at a time would mean the model is inaccurately evaluated at all wavelength pixels.  Instead we choose to evaluate all lines, but only on a subset $N_{\mathrm{batch}}$ of the total pixels $N_s$, so that the model can eventually converge to exact at those points.  All lines are allowed to update at each glimpse of a minibatch, but many lines with cores far from minibatch pixels will provide only weak information about how the loss scalar changes for their parameters.  The value of $N_{\mathrm{batch}}$ should be set just below the threshold at which the computation runs out of memory.  At present, this threshold is determined experimentally.  The indices of the wavelength points can either be set to be contiguous in blocks, or random with possibly large gaps between indices, or else with some other scheme tailored to the most informative wavelength regions.  We experimented with several techniques and settled on random sampling with replacement, meaning that some pixels get revisited more often than others, with some minibatches tailored only to the line cores.

@@

Another solution to the computation problem is to take advantage of the sparsity of the $\bm{\bar{V}}$ matrix: most of the Voigt profile's entries far from the line center are vanishingly close to zero.  Ideally we would only evaluate the model close to the line center position, some number of line cuts $N_{\mathrm{cut}}=\alpha \cdot \mathrm{FWHM}$ away.  These line cuts produce a speedup by a factor of $\frac{N_s}{N_{\mathrm{cut}}}$, which can exceed $100\times$ for wide bandwidth spectra. Furthermore, efficient algorithms for assembling and coalescing sparse matrices exist in \texttt{PyTorch}.  A few practical considerations make sparse matrices peskier to work with than their dense counterparts.  First, the $\alpha$ hyperparameter should be set large enough that the truncation effect is not seen for the deepest lines, or in post-processing steps such as line broadening.  The $\alpha$ should be preselected before the training process.  If the FWHM initialization was poor enough, the question may arise how autodiff training will cope with the poor performance of a model, leading to the prospect of strange and difficult to diagnose artifacts near spurious line-cut truncation artifacts.

@@

There are a few standardization strategies for applying the wing cuts.  Here we coerce all wing cuts to be exactly the same number of pixels, typically 6000 pixels for \texttt{PHOENIX}, with the middle pixel being at the line center position, and about 3000 pixels to the red and blue side of the line.  Most spectral lines only appreciably affect the signal in the central $\sim5-50$ pixels, so such a large wing cut may seem over-zealous.  We emphasize that the line cut serves as the fixed pixel locations where the line will be evaluated with post-processing steps such as radial velocity shifts and line broadening.  So these 6000 pixels must be capable of handling the fastest-moving and fastest-spinning stars you expect to see in your observed spectra of interest.

>>>>

\subsection{Seeking physical self-consistency and high predictive accuracy}
Ideally, we seek models that are both physically self-consistent and highly accurate when compared to data. We want tunable models that can be evaluated quickly for use in likelihood calculations. We want posterior inference to accurately explore correlations among parameters. We want unbiased, noise-free templates for use in cross-correlation spectroscopy. We want accuracy over a large spectral grasp: simultaneously wide bandwidth and high spectral resolution. We want high out-of-band prediction accuracy. Ideally, we also want precision and accuracy in the separation of affiliated telluric and instrumental signals.

@@

\subsection{This paper: towards semi-empirical models}
\begin{itemize}
 \item So in practice, tradeoffs between model flexibility and physical self-consistency
 \item Semi-empirical models may be a middle ground: informed from models, but revised with data
 \item The radiative transfer step is computationally expensive, so we want to avoid redoing that
 \item Instead focus on pre-computed models that have that wisdom baked-in
 \item We are essentially compressing pre-computed models into evaluable, interpretable models
 \item Transfer learning is one plausible approach: pre-train on pre-computed models, transfer to data
 \item Neural network frameworks make this possible with autodiff
 \item We use the autodiff and GPU acceleration, but not the neural architectures
 \item We do not use the neural architectures because we have a good understand for how a spectrum is generated
 \item Instead of tuning weights of a neural network, we tune the atomic and molecular properties of a spectrum
\end{itemize}

@@

We start with a high resolution pre-computed synthetic stellar model spectrum, $\mathsf{S}(\bm{\lambda})$. The procedure is largely agnostic to the exact details of how this spectrum was made. For the purposes of this paper, we will showcase examples from two well-known families of precomputed synthetic spectra: \texttt{PHOENIX} \citep{husser13} for stellar spectra $(T_{\mathrm{eff}}\in [2300, 10000]\;K)$ and \texttt{Sonora} \citep{2021ApJ...920...85M} for brown dwarf spectra $(T_{\mathrm{eff}}\in [1000, 2300]\;K)$. We place the following demands on the precomputed spectra. They should have sporadic regions of discernable continuum devoid of lines, and the continuum must vary smoothly in wavelength. The spectral lines or pseudo-lines should be resolved, and not sub-sampled. We suspect most stellar spectra meet these criteria, except for the coolest M dwarfs and brown dwarfs. The method can hypothetically handle spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed. We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars. A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

The choice of limiting the bandwidth to a region-of-interest around a single echelle spectrograph bandwidth stems from computational constraints. In principle, there is no fundamental limit to the bandwidth one could clone with the method presented here, up to and including the entire precomputed synthetic spectral model bandwidth. We adopt the exact native wavelength sampling with no smoothing or interpolation, yielding a wavelength vector $\bm{\lambda}_s$ with length $N_\mathrm{s}$ equal to the number of pixels within the extents of our region of interest including the buffers.

At this stage, we have the choice of whether to work in linear or log scale flux units. Adopting the log of the flux would ensure that the cloned model possesses only positive flux values, a desirable trait of any physical spectral model. We chose instead to stick with native linear flux scaling because of the availability of convenient approximations to astrophysically relevant line profiles in the linear scaling domain. We normalize the spectrum by its median value since we do not intend to apply these methods to absolutely flux calibrated spectra.

@@

We initialize the cloned model with a series of preprocessing steps. We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template. The resulting signal typically possesses a smooth continuum variation. An additional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly. Here, we identify the continuum regions by the most prominent peaks in the specturm separated by a distance of at least 5,000$-$90,000 pixels from adjacent peaks, yielding between 60 and 5 control points. We then fit an $n_{\mathrm{poly}} = 3^{rd}-15^{th}$ order polynomial to these peaks, evaluate the polynomial on the entire bandwidth $\mathsf{P}(\bm{\lambda}_s)$, and divide the spectrum by this trend. Visual inspection of this continuum-flattened spectrum typically possesses variations at the $<1\%$ level. Tuning the pixel distance and polynomial order can lower the undulations to $<0.05\%$ level for some input spectra. We found a pixel distance of 50,000 and $3^{rd}$ order polynomial worked across the range of effective temperatures for HPF's bandwidth. The IGRINS spectra had more significant continuum opacity and band heads to accommodate and therefore needed to capture finer-scale variation: 5,000 pixels and a $9^{th}$ order polynomial appeared adequate over most of the effective temperature range.

This high-pass filtering step should be set to capture the genuine spectral shape, without over-fitting broad line-wings such as those in deep Hydrogen and sodium lines. We have experimentally found it helpful to hand-tune models of exceptionally broad and non-standard line wings with an optional pre-processing step that depends on the wavelength range of interest. Here we sketch the approach for common broad lines found in the long-wavelength visible and near-infrared wavelength regions.

@@

We emphasize that a recreation of the unvarnished input spectrum can be obtained by multiplying the continuum-flattened signal by the ``perturbed black body'', $\mathsf{B}(\bm{\lambda}_s)\cdot \mathsf{P}(\bm{\lambda}_s)$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects. This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low-resolution spectra, or regions with molecular band heads. The ``perturbed black body'' continuum model contains $n_{\mathrm{poly}}+1$ fixed-but-possibly tunable lookup parameters, plus the fixed input $T_{\mathrm{eff}}$.

Next, we identify the spectral lines. We apply a local-minimum-finding algorithm on the spectrum by defining a prominence threshold $P_{rom} \in (0.005, 0.02)$. This threshold dictates the number of lines that will be modeled: a lower prominence finds more, weaker lines, and a larger prominence finds fewer, deeper lines. The prominence algorithm successfully finds lines that reside on top of broad line wings, or unresolved band heads provided that the individual lines exceed the prominence threshold in their local region. The number of lines $N_{\mathrm{lines}}$ depends on the bandwidth, prominence, and the intrinsic properties of the input spectrum, principally effective temperature and metallicity.

For this paper, we illustrate examples for two \'echelle spectrographs with particularly large spectral grasp: the Habitable Zone Planet Finder \citep[HPF,][]{2014SPIE.9147E..1GM} on the Hobby-Eberly Telescope at McDonald Observatory in Fort Davis, Texas; and the Immersion Grating Infrared Spectrograph \citep[IGRINS,][]{park14}
currently on the Gemini South Telescope on Cerro Pach\'on in Chile. The $R=55,000$ HPF has a native bandwidth of $8079-12785$~\AA, which we expand to $8049-12815$~\AA~including the edge buffers. IGRINS has two cameras for $H$ and $K$ band, with the combined spectrum spanning $14267-25217\;$\AA~ including the edge buffers and the region in-between the two cameras, all at a resolving power of $(R=45,000)$. The spectrograph acquisition, reduction, and post-processing steps yield data $\mathsf{D}(\bm{\lambda}_{d})$, where $\bm{\lambda}_{d}$ is the wavelength vector at the instrumental resolution and sampling of each instrument, generally much coarser than the resolution and sampling grid of the precomputed synthetic spectra. The data wavelength vector may also contain gaps between \'echelle orders, whereas the precomputed wavelength coordinates are usually contiguous. HPF may have up to $2048\times28=$57,344 pixels, and IGRINS has typically about 75,000 pixels, after common trimming of noisy edge pixels and unusable telluric regions.

@@

Figure \ref{fig_Nlines_vs_teff} shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold $P_{rom}$ for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS. We see between about 2,000 and 20,000 lines depending on the $T_{\mathrm{eff}}$ and $P_{rom}$. HPF and IGRINS have a comparable number of lines, and halving the prominence increases the number of lines by about $20-30\%$ in these ranges. The number of lines monotonically increases towards cooler effective temperatures.
The HPF-truncated spectra have $N_s=$335,849 native resolution samples, comparable to the IGRINS-truncated spectra, $N_s=$330,052.

@@

There does not exist a general-purpose, single-shot algorithm for obtaining the lineshape in the presence of overlapping spectral lines: where do the wings of one line begin and the wings of another adjacent line end? We, therefore, do not attempt to determine anything about the lineshape at this stage and instead assume that the lines resemble a Voigt profile, with a guess width about equally split between Lorentzian and Gaussian.

@@

where $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ is the normalized Voigt profile with Gaussian standard deviation $\sigma$, Lorentzian half-width $\gamma$, at line center position $\lambda_c$, for the $j^{th}$ spectral line. The amplitude $a$ is always expected to be positive for absorption lines and is equal to the area under the curve of the line since the Voigt profile is normalized. The optional Prefactor term $\mathsf{P}$ is fixed at $1$ by default but can be allowed to vary to fine-adjust imperfections leftover from the polynomial-based continuum flattening procedure.

\subsection{Voigt profile evaluation}
The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed-form using the Fadeeva function\authorcomment1{citation}. Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable. Here we adopt the pseudo-Voigt approximation described in \censor{some paper}\authorcomment1{look up reference to PseudoVoigt}.

Two sampled Voigt approaches could be feasible. Direct convolution of a sampled Gaussian and Lorentzian profile may be too costly and inexact if pixels are coarsely sampled. Analogously, the convolution can be achieved in the Fourier domain through the multiplication of the Fast Fourier Transforms of the Gaussian and Lorentzian profiles, followed by the subsequent Inverse Fourier Transforms. The main rationale for either of these two choices would be to anticipate and align with planned convolutions such as rotational broadening and instrumental kernels. We do not currently implement these sampled convolutions in \emph{blas\'e} due to their prohibitive computational cost.

\subsection{Goodness of fit metric}
The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes. Instead, we tune the parameters of the model, starting from these coarse initial values. This model has $N_{\mathrm{lines}}\times 3$ free parameters, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary. We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{M}$ but has no notion of per-pixel noise since the precomputed synthetic spectrum has no uncertainty:

@@

The optimizer updates the $a_j, \sigma_j, \gamma_j$ parameters by a small fraction of the Jacobian---called the learning rate LR---towards the direction that would improve the fit, for all $N_{\mathrm{lines}} \times 3$ parameters simultaneously. The Jacobian is calculated behind the scenes with automatic differentiation\authorcomment1{Citation} implemented as the so-called backpropagation algorithm or simply ``backprop''. We choose the \texttt{PyTorch} framework that computes these Jacobians efficiently for all of the mathematical primitives in our \texttt{blas\'e} implementation.

In principle, we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch. Further, an autodiff-aware implementation of the \emph{celerit\'e} algorithm\authorcomment1{add DFM citation} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting. In practice, PyTorch also offers hardware acceleration (on GPU or TPU)\authorcomment1{citation needed}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks. The GPyTorch framework\authorcomment1{add citation} may offer a workaround to the absence of a \emph{celerit\'e} implementation. Still, the performance of an analytic Jacobian implementation over the autodiff route presented here offers an interesting avenue of future research.

\subsection{GPU and Autodiff specific considerations}
We make a few tweaks to the implementation for numerical purposes. First, we want all the parameters to be positive, forbidding negative amplitudes and negative widths. We, therefore, tune the natural log of the parameters and exponentiate them before inclusion in Equation \ref{equation1}. Second, we found through iterative experimentation that the initialization amplitudes and widths were systematically shifted from the optimized values. We built-in scalar tweaks to the initialization amplitudes, Gaussian width, and Lorentzian width, which dramatically accelerated the optimization step.

@@

Equation \ref{equation1} performs a type of tensor contraction, turning a $N_{\mathrm{lines}}\times N_{s}$ matrix into a $1\times N_{s}$ vector. The number of Floating Point Operations (FLOPS) scales with the number of entries in this matrix, with a scalar prefactor for the cost of evaluating a single line profile at a single wavelength point. Efficient GPU algorithms exist for tensor contractions such as this product along an axis of a matrix, allowing this computation to proceed quickly on modern machines\authorcomment1{add citation?}. In particular, the proprietary CUDA architecture for NVIDIA\textsuperscript{\tiny\textregistered} GPUs contains Tensor cores with specialized matrix math, although the algorithms place restrictions on floating precision depending on the particular hardware and software available. The chief bottleneck occurs when the storage of the $\bm{\bar{V}}$ matrix exceeds the available RAM of a GPU or CPU: the computation will fail with an ``Out of Memory'' exception. Modern NVIDIA GPUs have $8-40$ GB of RAM. The memory bottleneck is even more pernicious than mere storage of the $\bm{\bar{V}}$ matrix since the CPU/GPU also has to store the computation graph for autodiff. The computation graph takes up more memory than the matrix itself, so it is generally not possible to evaluate Equation \ref{equation1} in its entirety in one-fell-swoop. A remedy is needed.

@@

Minibatches offer a natural solution to the line-by-line computational bottleneck while also providing optimization performance improvements. Minibatches act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent\authorcomment1{citation}. The Machine Learning/Artificial Intelligence (ML/AI) community employs minibatches to handle similar situations, in which the dataset volume exceeds the available GPU RAM. Our configuration departs slightly from the typical ML/AI situation since their models are usually ``small'' relative to their datasets. Here our model is ``big'' relative to our dataset.

Here we choose to assemble and evaluate only a portion of the matrix at a time, in minibatches. The choice to evaluate only a portion of lines at a time would mean the model is inaccurately evaluated at all wavelength pixels. Instead, we choose to evaluate all lines, but only on a subset $N_{\mathrm{batch}}$ of the total pixels $N_s$, so that the model can eventually converge to exact at those points. All lines are allowed to update at each glimpse of a minibatch, but many lines with cores far from minibatch pixels will provide only weak information about how the loss scalar changes for their parameters. The value of $N_{\mathrm{batch}}$ should be set just below the threshold at which the computation runs out of memory. At present, this threshold is determined experimentally. The indices of the wavelength points can either be set to be contiguous in blocks, or random with possibly large gaps between indices, or else with some other scheme tailored to the most informative wavelength regions. We experimented with several techniques and settled on random sampling with replacement, meaning that some pixels get revisited more often than others, with some minibatches tailored only to the line cores.

@@

Another solution to the computation problem is to take advantage of the sparsity of the $\bm{\bar{V}}$ matrix: most of the Voigt profile's entries far from the line center are vanishingly close to zero. Ideally, we would only evaluate the model close to the line center position, some number of line cuts $N_{\mathrm{cut}}=\alpha \cdot \mathrm{FWHM}$ away. These line cuts produce a speedup by a factor of $\frac{N_s}{N_{\mathrm{cut}}}$, which can exceed $100\times$ for wide bandwidth spectra. Furthermore, efficient algorithms for assembling and coalescing sparse matrices exist in \texttt{PyTorch}. A few practical considerations make sparse matrices peskier to work with than their dense counterparts. First, the $\alpha$ hyperparameter should be set large enough that the truncation effect is not seen for the deepest lines, or in post-processing steps such as line broadening. The $\alpha$ should be preselected before the training process. If the FWHM initialization was poor enough, the question may arise how autodiff training will cope with the poor performance of a model, leading to the prospect of strange and difficult-to-diagnose artifacts near spurious line-cut truncation artifacts.

@@

There are a few standardization strategies for applying wing cuts. Here we coerce all wing cuts to be the same number of pixels, typically 6000 pixels for \texttt{PHOENIX}, with the middle pixel being at the line center position, and about 3000 pixels to the red and blue side of the line. Most spectral lines only appreciably affect the signal in the central $\sim5-50$ pixels, so such a large wing cut may seem over-zealous. We emphasize that the line cut serves as the fixed pixel locations where the line will be evaluated with post-processing steps such as radial velocity shifts and line broadening. So these 6000 pixels must be capable of handling the fastest-moving and fastest-spinning stars you expect to see in your observed spectra of interest.

The PHOENIX spectra double their pixel sampling shortward of $\lambda = 1 \mu$m, so 6000 pixels in the visible is about 30 Angstroms, but 6000 pixels in the infrared is about 60 Angstroms. We \emph{could have} demanded that the wing cut be exactly 30 $\AA$, or some other heuristic, but that choice would result in so-called \emph{jagged} or \emph{ragged} arrays: collections of 1D arrays that differ in their lengths. Some lines would have 5999 pixels, some would have 6000, and others would have 2999 or 3000 pixels, depending whether they occur to the red or blue side of the $1 \;\mu$m kink in sampling. Most efficient CPU and GPU algorithms are not equipped to deal with ragged arrays, therefore increasing the computation time. Instead, the standardization choice of exactly 6000 pixels means we can use standard matrix or tensor arithmetic, which somewhat counterintuitively is faster while handling more pixels for all spectral lines: 6000 pixels everywhere is faster than 6000 pixels for some and 2999 for other lines.
